Viscous Transport in Eroding Porous Media

### 2019-10-12 docs/transport/erosion_transport.tex
NM
done up to section 6

\begin{abstract} 
  Transport of viscous fluid through porous media is a direct
  consequence of the pore structure. Here we investigate transport
  through a specific class of two-dimensional porous geometries, namely
  those formed by fluid-mechanical erosion.  We investigate the
  tortuosity and dispersion by analyzing the first two statistical
  moments of tracer trajectories. For most initial configurations,
  tortuosity decreases in time as a result of erosion increasing the
  porosity.  However, we find that tortuosity can also increase
  transiently in certain cases.  The porosity-tortuosity relationships
  that result from our simulations are compared against models available
  in the literature.  Asymptotic dispersion rates are also strongly
  affected by the erosion process, as well as by the number and
  distribution of the eroding bodies. Finally, we analyze the pore size
  distribution of an eroding geometry. The simulations are performed by
  combining a high-fidelity boundary integral equation solver for the
  fluid equations, a second-order stable time stepping method to
  simulate erosion, and new numerical methods to stably and accurately
  resolve nearly-touching eroded bodies and particle trajectories near
  the eroding bodies.

@@

To characterize transport through the resulting configurations, particle
trajectories must be computed. Depending on the application, microscale
transport can be modelled as pure advection~\citep{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019},
advection-diffusion~\citep{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\citep{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume trajectories to be governed by pure advection
(ie.~no diffusion), so the particle trajectories are identical to the
streamlines. We compute trajectories $\ss(t)$ that are initialized at
$\ss_0$ by solving the advection equation
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Since there is no stiff diffusive
term in~\eqref{eqn:tracers}, we use a fourth-order explicit Runge-Kutta
time stepping method.  In addition, the Barycentric quadrature rule is
used to accurately compute trajectories that are close to an eroding
grain. We note that an alternative approach to compute trajectories is
to form a distribution of velocities $\uu$ and use a continuous time
random walk~\citep{mon-wei1965, den-cor-sch-ber2004, leb-den-car2008,
ber-cor-den-sch2006}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Contributions}
Our first contribution is extending the Barycentric quadrature rule
of~\citet{bar-wu-vee2015} to accurately compute velocity gradients
associated with Stokes flow. This new quadrature rule permits accurate
and stable simulations of erosion, and, moreover, enables visualization
of the vorticity field throughout.

Our second contribution is computing and analyzing trajectories in
eroded geometries.  These trajectories are formed by combining the
Barycentric quadrature rule and a fourth-order Runge-Kutta method. Then,
they are used to compute tortuosities and dispersion rates of eroded
geometries. Finally, we examine the effect of erosion on the pore size
distribution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Limitations}
As in our previous work~\citep{qua-moo2018}, we consider two-dimensional
immobile eroding grains. However, this work does not limit the proximity
of  grains. Compared to our previous work, we are able to simulate much
denser suspensions with a comparable number of discretization points and
without upsampling.  While we do investigate transport in this work, we
only consider passive particles that are advected by the flow.  To model
other physical processes, such as contaminant transport, we plan to
simulate a finite Peclet number advection-diffusion equation in eroded
geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Outline of the Paper}
In Section~\ref{sec:formulation}, we summarize the erosion model that is
described in more detail in previous work~\citep{qua-moo2018}.  In
Section~\ref{sec:DLP}, we recast all the governing equations as layer
potentials defined in both $\RR^2$ and in $\CC$.
Section~\ref{sec:transport} describes measures for characterizing the
geometry and transport. Section~\ref{sec:method} describes the numerical
methods, with special attention paid to the new quadrature method for
computing the shear stress and the vorticity.  Section~\ref{sec:results}
presents numerical examples for a variety of dense packings of bodies.
Finally, concluding remarks are made in Section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  We only
briefly summarize the model, and a more detailed description can be
found in previous work~\citep{qua-moo2018}.  We consider flows inside a
confined geometry $\Omega$ that contains $M$ eroding bodies with
boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$.  The boundary of the
fluid domain is $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup
\gamma_M$, where $\Gamma$ is the outer boundary, taken to be a slightly
smoothed version of the boundary of $[-3,3] \times [-1,1]$.  We only
place eroding bodies in $[-1,1] \times [-1,1]$ to create a buffer region
for the flow profile imposed at the inlet to transition to the more
complex flow intervening between the bodies. On each eroding body, $\nn$
is the normal vector pointing into the body, and $\ss$ is the unit
tangent vector pointing in the counterclockwise direction.  The normal
vector on $\Gamma$ is the outward unit normal.  Neglecting inertial
forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the eroding bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here, $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, $\Vn$ is the normal velocity
of $\gamma$, and
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
is the shear stress on $\gamma$. We simulate erosion by alternating
between solving the fluid equations and advancing the erodible
boundaries. The strength of $\UU$ is adjusted at each time step to
achieve a constant pressure drop across the channel.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Boundary Integral Equation Formulation}
\label{sec:DLP}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Since we require an accurate solution of the incompressible Stokes
equations in complex two-dimensional geometries, we reformulate the
equations as a BIE.  This has the advantage that only the
one-dimensional boundary of the domain must be discretized, and, with
appropriate quadrature formulas and fast summation methods, the result
is a high-fidelity numerical simulation with near-optimal computational
complexity.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cauchy Integral Representation of the Double-Layer
Potential}
\label{sec:DLPcomplex}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In
equations~\eqref{eqn:velocityDLP},~\eqref{eqn:deformationDLP},~\eqref{eqn:pressureDLP},
and~\eqref{eqn:vorticityDLP}, the fluid velocity, deformation tensor,
pressure, and vorticity are all written as layer potentials in $\RR^2$.
However, the quadrature method we introduce in Section~\ref{sec:method}
requires a complex-valued representation of the layer potentials for the
velocity, deformation tensor, and vorticity. The first step to form a
complex representation is to write the Laplace double-layer potential as
the complex integral
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{2\upi} \int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2}\eeta(\yy)\, ds_\yy = \Real (v(x)),
\end{align}
where
\begin{align}
  v(x) = \frac{1}{2\upi i} \int_{\bd\Omega}
    \frac{\eta(y)}{x - y} \, dy, \quad x \in \Omega.
  \label{eqn:laplaceComplex}
\end{align}
Here $x = x_1 + i x_2,y = y_1 + i y_2 \in \CC$ are the complex
counterparts of $\xx = (x_1,x_2),\yy = (y_1,y_2) \in \RR^2$, and $\eta =
\eta_1 + i \eta_2$ is the complex counterpart of $\eeta =
(\eta_1,\eta_2)$. Therefore, depending on the formulation of the layer
potential, $\Omega$ is being interpreted as a subset of $\RR^2$ or
$\CC$. Equation~\eqref{eqn:laplaceComplex} is converted to a Cauchy
integral by first finding the boundary data of $v$.  Assuming that
$\Omega$ is a simply-connected interior domain, the boundary data of $v$
satisfies the Sokhotski-Plemelj jump relation
\begin{align}
  \label{eqn:SPrelation}
  v(x) = - \frac{1}{2} \eta(x) + \frac{1}{2\upi i} \int_{\bd\Omega}
    \frac{\eta(y)}{x-y}\, dy, \quad x \in \Omega.
\end{align}
For exterior domains, the jump term changes from $-1/2$ to $1/2$, and
for multiply-connected domains, such as a porous media, $\bd\Omega$ is
decomposed into its different connected components and the appropriate
jump relation is applied.  Having computed the boundary data of the
holomorphic function $v$, by the Cauchy integral theorem we have
\begin{subequations}
  \begin{alignat}{3}
  \label{eqn:cauchy}
  v(x) &= \frac{1}{2\upi i}\int_{\bd\Omega} 
    \frac{v(y)}{y-x} \,dy, \\
  v'(x) &= \frac{1}{2\upi i} \int_{\bd\Omega}
    \frac{v(y)}{(y-x)^2} \, dy, \\
  v''(x) &= \frac{1}{\upi i} \int_{\bd\Omega}
    \frac{v(y)}{(y-x)^3} \, dy,
  \end{alignat}
\end{subequations}
for $x \in \Omega$.  Since $v(x)$ depends on the complex-valued density
function $\eta$, we use the notation $v[\eta](x)$ for the holomorphic
function defined in equation~\eqref{eqn:laplaceComplex}, and its first
two derivatives are written as $v'[\eta](x)$ and $v''[\eta](x)$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cauchy Integral Representation for the Gradient of the
Double-Layer Potential}
\label{sec:gradDLPcomplex}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computing the shear stress and vorticity requires a complex-valued layer
potential representation of the velocity gradient.  The deformation
tensor at $x \in \Omega$ is found by computing the derivatives of the
expressions for $u_1$ and $u_2$ in equation~\eqref{eqn:cauchyVelocity}  
\begin{equation}
\label{eqn:cauchyGradient}
  \begin{aligned}
    \pderiv{u_1}{x_1} &= +\Real (v'[\psi_1](x)) + 
    \Real (v''[y\cdot\eta](x)) - \Real (v'[\eta_1](x)) \\
    &- x_1\Real (v''[\eta_1](x)) - x_2\Real (v''[\eta_2](x)), \\
    \pderiv{u_1}{x_2} &= - \Imag (v'[\psi_1](x)) - 
    \Imag (v''[y\cdot\eta](x)) + x_1\Imag (v''[\eta_1](x)) \\
    &- \Real (v'[\eta_2](x)) + x_2\Imag (v''[\eta_2](x)), \\
    \pderiv{u_2}{x_1} &= +\Real (v'[\psi_2](x)) - 
    \Imag (v''[y\cdot\eta](x)) + \Imag (v'[\eta_1](x))  \\
    &+ x_1\Imag (v''[\eta_1](x)) + x_2\Imag (v''[\eta_2](x)), \\
    \pderiv{u_2}{x_2} &= -\Imag (v'[\psi_2](x)) - 
    \Real (v''[y\cdot\eta](x)) + x_1\Real (v''[\eta_1](x)) \\
    &+ \Imag (v'[\eta_2](x)) + x_2\Real (v''[\eta_2](x)).
  \end{aligned}
\end{equation}
The same expressions are used to compute the deformation tensor for $x
\in \bd\Omega$, except that the jump
condition~\eqref{eqn:deformationJump} is included.  Finally, to compute
the shear stress, the deformation tensor on $\bd\Omega$ is applied to
the normal and tangent vectors as in equation~\eqref{eqn:shearStress}.
The velocity gradient is also used to compute the vorticity in the fluid
bulk.  For $x \in \Omega$, the Cauchy integral representation of the
vorticity at $x \in \Omega$ is
\begin{equation}
  \begin{aligned}
    \omega(x) = \pderiv{u_2}{x_1} - \pderiv{u_1}{x_2} &= 
      \Real (v'[\psi_2](x)) + \Imag (v'[\psi_1](x))  \\
      &+ \Real (v'[\eta_2](x))+ \Imag (v'[\eta_1](x)).
  \end{aligned}
\end{equation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Anomalous Dispersion}
\label{sec:dispersion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The spreading of fluid in a porous media is often characterized in terms
of anomalous dispersion~\citep{kla-rad-sok2008, den-cor-sch-ber2004}.
The anomalous dispersion rate depends on the porosity and
permeability~\citep{koc-bra1988}, but is also affected by the
distribution and shape of the grains. We calculate the anomalous
dispersion rates by analyzing the streamlines governed by
equation~\eqref{eqn:tracers} in eroded geometries. Given a set of $N_p$
trajectories, we define $\lambda_j(t)$ to be the arclength of the
trajectory
\begin{align}
  \lambda_j(t) = \int_{0}^t \|\ss'_j(\tilde{t})\|\, d\tilde{t}, 
    \quad j=1,\ldots,N_p.
\end{align}
Then, the first and second ensemble moments are
\begin{align}
  \label{eqn:moments}
  \langle \lambda \rangle (t) = 
    \frac{1}{N_p} \sum_{j=1}^{N_p} \lambda_j(t), \quad 
    \sigma_\lambda^2(t) = \frac{1}{N_p} 
    \left[\lambda_j(t) - \langle \lambda \rangle(t) \right]^2,
\end{align}
and $\sigma_\lambda$ characterizes particle dispersion.  At early times,
the particles have not explored much of the geometry, and we expect a
ballistic motion $\sigma_\lambda \sim t$.  However, as the particles
pass the grains, their trajectories are altered, and we expect that
$\sigma_\lambda \sim t^\alpha$, with $\alpha \in (0.5,1)$, indicating
that the flow is super-dispersive.

@@

We consider trajectories in an open channel so that they can be compared
to trajectories in eroded geometries with large permeability.  The
trajectory originating at $(s_1(0),s_2(0))$ is
\begin{align}
  s_1(t) = (1-s_2(0)^2)t + s_1(0), \quad
  s_2(t) = s_2(0).
\end{align}
Averaging over $s_2(0) \in [-1,1]$, the first and second ensemble
moments are
\begin{align}
  \langle \lambda \rangle (t) = \frac{2}{3}t, \quad 
    \sigma_\lambda^2(t) = \frac{4}{45}t^2.
\end{align}
In Section~\ref{sec:results}, we compare the second moment in an open
channel with the trajectories in an eroded geometry.

>>>>

\begin{abstract} 
  Transport of viscous fluid through porous media is a direct
  consequence of the pore structure. Here we investigate transport
  through a specific class of two-dimensional porous geometries, namely
  those formed by fluid-mechanical erosion.  We investigate the
  tortuosity and dispersion by analyzing the first two statistical
  moments of tracer trajectories. For most initial configurations,
  tortuosity decreases in time as a result of erosion increasing the
  porosity.  However, we find that tortuosity can also increase
  transiently in certain cases.  The porosity-tortuosity relationships
  that result from our simulations are compared with models available in
  the literature.  Asymptotic dispersion rates are also strongly
  affected by the erosion process, as well as by the number and
  distribution of the eroding bodies. Finally, we analyze the pore size
  distribution of an eroding geometry. The simulations are performed by
  combining a high-fidelity boundary integral equation solver for the
  fluid equations, a second-order stable time stepping method to
  simulate erosion, and new numerical methods to stably and accurately
  resolve nearly-touching eroded bodies and particle trajectories near
  the eroding bodies.

@@

To characterize transport through the resulting configurations, particle
trajectories must be computed. Depending on the application, microscale
transport can be modelled as pure advection~\citep{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019},
advection-diffusion~\citep{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\citep{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume trajectories to be governed by pure advection
(ie.~no diffusion), so the particle trajectories are identical to the
streamlines. We compute trajectories $\ss(t)$ that are initialized at
$\ss_0$ by solving the advection equation
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Since there is no stiff diffusive
term, we solve~\eqref{eqn:tracers} with a fourth-order explicit
Runge-Kutta time stepping method.  In addition, the Barycentric
quadrature rule is used to accurately compute trajectories that are
close to an eroding grain. 
%We note that an alternative approach to
%compute trajectories is to form a distribution of velocities $\uu$ and
%use a continuous time random walk~\citep{mon-wei1965,
%den-cor-sch-ber2004, leb-den-car2008, ber-cor-den-sch2006}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Contributions}
%Our first contribution is extending the Barycentric quadrature rule
%of~\citet{bar-wu-vee2015} to accurately compute velocity gradients
%associated with Stokes flow. This new quadrature rule permits accurate
%and stable simulations of erosion, and, moreover, enables visualization
%of the vorticity field throughout.
%
%Our second contribution is computing and analyzing trajectories in
%eroded geometries.  These trajectories are formed by combining the
%Barycentric quadrature rule and a fourth-order Runge-Kutta method. Then,
%they are used to compute tortuosities and dispersion rates of eroded
%geometries. Finally, we examine the effect of erosion on the pore size
%distribution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Limitations}
%As in our previous work~\citep{qua-moo2018}, we consider two-dimensional
%immobile eroding grains. However, this work does not limit the proximity
%of  grains. Compared to our previous work, we are able to simulate much
%denser suspensions with a comparable number of discretization points and
%without upsampling.  While we do investigate transport in this work, we
%only consider passive particles that are advected by the flow.  To model
%other physical processes, such as contaminant transport, we plan to
%simulate a finite Peclet number advection-diffusion equation in eroded
%geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Outline of the Paper}
This paper is organized as follows. In section~\ref{sec:formulation}, we
summarize the erosion model that is described in more detail in previous
work~\citep{qua-moo2018}.  In section~\ref{sec:DLP}, we recast all the
governing equations as layer potentials defined in both $\RR^2$ and in
$\CC$.  Section~\ref{sec:transport} describes measures for
characterizing the geometry and transport. Section~\ref{sec:method}
describes the numerical methods, with special attention paid to the new
quadrature method for computing the shear stress and the vorticity.
Section~\ref{sec:results} presents numerical examples for a variety of
dense packings of bodies.  Finally, concluding remarks are made in
section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  We only
briefly summarize the model, and a more detailed description can be
found in previous work~\citep{qua-moo2018}.  We consider flows inside a
confined geometry $\Omega$ that contains $M$ eroding bodies with
boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$.  The boundary of the
fluid domain is $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup
\gamma_M$, where $\Gamma$ is the outer boundary, taken to be a slightly
smoothed version of the boundary of $[-3,3] \times [-1,1]$.  We only
place eroding bodies in $[-1,1] \times [-1,1]$ to create a buffer region
for the flow profile imposed at the inlet to transition to the more
complex flow intervening between the bodies. On each eroding body, $\nn$
is the normal vector pointing into the body, and $\ss$ is the unit
tangent vector pointing in the counterclockwise direction.  The normal
vector on $\Gamma$ is the outward unit normal.  Neglecting inertial
forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the eroding bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here, $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, $\Vn$ is the normal velocity
of $\gamma$, and
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
is the shear stress on $\gamma$. We simulate erosion by alternating
between solving the fluid equations and advancing the eroding grains.
The strength of $\UU$ is adjusted at each time step to achieve a
constant pressure drop across the channel.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Boundary Integral Equation Formulation}
\label{sec:DLP}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Since we require accurate solutions of the incompressible Stokes
equations in complex two-dimensional geometries, we reformulate the
equations as a BIE.  This has the advantage that only the
one-dimensional boundary of the domain must be discretized, and, with
appropriate quadrature formulas and fast summation methods, the result
is a high-fidelity numerical simulation with near-optimal computational
complexity.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cauchy Integral Representation of the Double-Layer
Potential}
\label{sec:DLPcomplex}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In
equations~\eqref{eqn:velocityDLP},~\eqref{eqn:deformationDLP},~\eqref{eqn:pressureDLP},
and~\eqref{eqn:vorticityDLP}, the fluid velocity, deformation tensor,
pressure, and vorticity are all written as layer potentials in $\RR^2$.
However, the quadrature method we introduce in section~\ref{sec:method}
requires a complex-valued representation of the layer potentials for the
velocity, deformation tensor, and vorticity. The first step to form a
complex representation is to write the Laplace double-layer potential as
the complex integral
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{2\upi} \int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2}\eeta(\yy)\, ds_\yy = \Real (v(x)),
\end{align}
where
\begin{align}
  v(x) = \frac{1}{2\upi i} \int_{\bd\Omega}
    \frac{\eta(y)}{x - y} \, dy, \quad x \in \Omega.
  \label{eqn:laplaceComplex}
\end{align}
Here $x = x_1 + i x_2,y = y_1 + i y_2 \in \CC$ are the complex
counterparts of $\xx = (x_1,x_2),\yy = (y_1,y_2) \in \RR^2$, and $\eta =
\eta_1 + i \eta_2$ is the complex counterpart of $\eeta =
(\eta_1,\eta_2)$. Therefore, depending on the formulation of the layer
potential, $\Omega$ is being interpreted as a subset of $\RR^2$ or
$\CC$. Equation~\eqref{eqn:laplaceComplex} is converted to a Cauchy
integral by first finding the boundary data of $v$.  Assuming that
$\Omega$ is a simply-connected interior domain, the boundary data of $v$
satisfies the Sokhotski-Plemelj jump relation
\begin{align}
  \label{eqn:SPrelation}
  v(x) = - \frac{1}{2} \eta(x) + \frac{1}{2\upi i} \int_{\bd\Omega}
    \frac{\eta(y)}{x-y}\, dy, \quad x \in \Omega.
\end{align}
For exterior domains, the jump term changes from $-1/2$ to $1/2$, and
for multiply-connected domains, such as a porous media, $\bd\Omega$ is
decomposed into its different connected components and the appropriate
jump relation is applied.  Having computed the boundary data of the
holomorphic function $v$, by the Cauchy integral theorem we have
\begin{subequations}
  \label{eqn:cauchy}
  \begin{alignat}{3}
  v(x) &= \frac{1}{2\upi i}\int_{\bd\Omega} 
    \frac{v(y)}{y-x} \,dy, \\
  v'(x) &= \frac{1}{2\upi i} \int_{\bd\Omega}
    \frac{v(y)}{(y-x)^2} \, dy, \\
  v''(x) &= \frac{1}{\upi i} \int_{\bd\Omega}
    \frac{v(y)}{(y-x)^3} \, dy,
  \end{alignat}
\end{subequations}
for $x \in \Omega$.  Since $v(x)$ depends on the complex-valued density
function $\eta$, we use the notation $v[\eta](x)$ for the holomorphic
function defined in equation~\eqref{eqn:laplaceComplex}, and its first
two derivatives are written as $v'[\eta](x)$ and $v''[\eta](x)$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cauchy Integral Representation for the Gradient of the
Double-Layer Potential}
\label{sec:gradDLPcomplex}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computing the shear stress and vorticity requires a complex-valued layer
potential representation of the velocity gradient.  The deformation
tensor at $x \in \Omega$ is found by computing the derivatives of the
expressions for $u_1$ and $u_2$ in equation~\eqref{eqn:cauchyVelocity}  
\begin{equation}
\label{eqn:cauchyGradient}
  \begin{aligned}
    \pderiv{u_1}{x_1} &= +\Real (v'[\psi_1](x)) + 
    \Real (v''[y\cdot\eta](x)) - \Real (v'[\eta_1](x)) \\
    &- x_1\Real (v''[\eta_1](x)) - x_2\Real (v''[\eta_2](x)), \\
    \pderiv{u_1}{x_2} &= - \Imag (v'[\psi_1](x)) - 
    \Imag (v''[y\cdot\eta](x)) + x_1\Imag (v''[\eta_1](x)) \\
    &- \Real (v'[\eta_2](x)) + x_2\Imag (v''[\eta_2](x)), \\
    \pderiv{u_2}{x_1} &= +\Real (v'[\psi_2](x)) - 
    \Imag (v''[y\cdot\eta](x)) + \Imag (v'[\eta_1](x))  \\
    &+ x_1\Imag (v''[\eta_1](x)) + x_2\Imag (v''[\eta_2](x)), \\
    \pderiv{u_2}{x_2} &= -\Imag (v'[\psi_2](x)) - 
    \Real (v''[y\cdot\eta](x)) + x_1\Real (v''[\eta_1](x)) \\
    &+ \Imag (v'[\eta_2](x)) + x_2\Real (v''[\eta_2](x)).
  \end{aligned}
\end{equation}
The same expressions are used to compute the deformation tensor for $x
\in \bd\Omega$, except that the jump
condition~\eqref{eqn:deformationJump} is included.  Finally, to compute
the shear stress, the deformation tensor on $\bd\Omega$ is applied to
the normal and tangent vectors as in equation~\eqref{eqn:shearStress}.
The velocity gradient is also used to compute the vorticity in the fluid
bulk.  For $x \in \Omega$, the Cauchy integral representation of the
vorticity at $x \in \Omega$ is
\begin{align}
  \omega(x) = 
    \Real (v'[\psi_2](x)) + \Imag (v'[\psi_1](x))+ 
    \Real (v'[\eta_2](x))+ \Imag (v'[\eta_1](x)).
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Anomalous Dispersion}
\label{sec:dispersion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The spreading of fluid in a porous media is often characterized in terms
of anomalous dispersion~\citep{kla-rad-sok2008, den-cor-sch-ber2004}.
The anomalous dispersion rate depends on the porosity and
permeability~\citep{koc-bra1988}, but is also affected by the
distribution and shape of the grains. We calculate the anomalous
dispersion rates by analyzing the streamlines governed by
equation~\eqref{eqn:tracers} in eroded geometries. Given a set of $N_p$
trajectories, we define $\lambda_j(t)$ to be the arclength of the
trajectory
\begin{align}
  \lambda_j(t) = \int_{0}^t \|\ss'_j(\tilde{t})\|\, d\tilde{t}, 
    \quad j=1,\ldots,N_p.
\end{align}
Then, the first and second ensemble moments are
\begin{align}
  \label{eqn:moments}
  \langle \lambda \rangle (t) = 
    \frac{1}{N_p} \sum_{j=1}^{N_p} \lambda_j(t), \quad 
    \sigma_\lambda^2(t) = \frac{1}{N_p} \sum_{j=1}^{N_p}
    \left[\lambda_j(t) - \langle \lambda \rangle(t) \right]^2,
\end{align}
and $\sigma_\lambda$ characterizes particle dispersion.  At early times,
the particles have not explored much of the geometry, and we expect a
ballistic motion $\sigma_\lambda \sim t$.  However, as the particles
pass the grains, their trajectories are altered, and we expect that
$\sigma_\lambda \sim t^\alpha$, with $\alpha \in (0.5,1)$, indicating
that the flow is super-dispersive.

@@

We consider trajectories in an open channel so that they can be compared
to trajectories in eroded geometries with large permeability.  The
trajectory originating at $(s_1(0),s_2(0))$ is
\begin{align}
  s_1(t) = (1-s_2(0)^2)t + s_1(0), \quad
  s_2(t) = s_2(0).
\end{align}
Averaging over $s_2(0) \in [-1,1]$, the first and second ensemble
moments are
\begin{align}
  \langle \lambda \rangle (t) = \frac{2}{3}t, \quad 
    \sigma_\lambda^2(t) = \frac{4}{45}t^2.
\end{align}
In section~\ref{sec:results}, we compare the second moment in an open
channel with the trajectories in an eroded geometry.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The tortuosity is a dimensionless number that quantifies the amount of
twisting of streamlines. Unlike the dispersion calculations, we do not
use reinsertion to form long trajectories.  The local tortuosity is
\begin{align}
  \tau(y_0) = \frac{\lambda(y_0)}{d}.
  \label{eqn:localTort}
\end{align}
Here the streamline originates on the inlet cross-section $x=x_0$ at
$(x_0,y_0)$, and its arclength, $\lambda(y_0)$, is calculated until the
streamline passes the parallel outlet cross-section $x = x_0 + d$.  In
this work, we consider streamlines originating at $x=-1$ and terminating
at $x=1$, so $d=2$.  However, other choices for the terminal point when
computing the tortuosity are sometime used~\citep{dud-koz-mat2011}.  The
hydraulic tortuosity is defined by taking the average over all points on
the inlet cross-section
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\int_{S}u_1(x_0,y_0)\lambda(y_0)\,dy_0}
  {\displaystyle\int_{S}u_1(x_0,y_0)\,dy_0},
  \label{eqn:tortuosity1}
\end{align} 
where $S$ is the inlet cross-section $x = -1$, $u_1(y_0)$ is the
$x$-component of the velocity at the initial point of the streamline,
and $d=2$ is the distance between the inlet and outlet.  Note that $T
\geq 1$, and $T=1$ only if no grains are present.
